Boosting Higgs Pair Production in the $b\bar{b}b\bar{b}$ Final State with Multivariate Techniques

### 2015-12-15 paper/sec-intro.tex

The measurement of double Higgs production will be one of the central
physics goals of the LHC program in its recently started high-energy
phase, as well as for its future high-luminosity upgrade (HL-LHC),
which aims to accumulate a total integrated
luminosity of 3 ab$^{-1}$~\cite{ATLAS:2013hta,CMS:2013xfa}.
%
Higgs pair production is directly sensitive to the
Higgs trilinear coupling $\lambda$~\cite{baglio} and 
provides crucial
information on the electroweak symmetry breaking mechanism.
%
It also probes the underlying strength of the Higgs interactions
at high energies, and can be used to test the composite nature of the 
Higgs boson~\cite{Giudice:2007fh,Contino:2010mh}.
%
While Standard Model (SM) cross-sections are small,
many Beyond the SM (BSM)
scenarios predict enhanced rates for double Higgs production, and
thus searches for this process have already been performed by ATLAS and CMS with Run I data~\cite{Aad:2015xja,Aad:2015uka,Aad:2014yja,Khachatryan:2015yea,Chatrchyan:2011wt}
and will continue at Run II.
%
In addition, Higgs pair production will also be of
crucial relevance at
any future high-energy 
collider, either at a 100 TeV circular machine~\cite{Arkani-Hamed:2015vfh,Barr:2014sga,Papaefstathiou:2015iba,Azatov:2015oxa} or at
a linear or circular electron-positron collider~\cite{Contino:2013gna}.

Analogously to single Higgs production~\cite{Dittmaier:2012vm}, 
in the SM the dominant mechanism for the production of a pair of
Higgs bosons at the LHC is 
gluon fusion (see Refs.~\cite{baglio,Frederix:2014hta} and
references therein).
%
For a center-of-mass energy $\sqrt{s} = 14\,$TeV, the
next-to-next-to-leading order (NNLO)
total cross section is approximately $40\,$fb~\cite{deFlorian:2013jea},
which is increased by a further few percent once
next-to-next-to-leading logarithmic
(NNLL) corrections
are accounted for~\cite{deFlorian:2015moa}.
%
Feasibility studies in the case of a SM-like Higgs boson
in the gluon-fusion channel
at the LHC have been performed for different final states, including
$b\bar b\gamma\gamma$~\cite{Baur:2003gp,Barger:2013jfa},
$b\bar{b}\tau^+\tau^-$~\cite{Baur:2003gpa,Barr:2013tda,Dolan:2012rv,Dolan:2013rja},
$b\bar{b}W^+W^-$~\cite{Dolan:2012rv,Papaefstathiou:2012qe} and
$b\bar{b}b\bar{b}$~\cite{Baur:2003gpa,Dolan:2012rv,Wardrope:2014kya,deLima:2014dta,Barger:2013jfa}.
%
While these studies differ in their quantitative conclusions,
a consistent picture emerges 
that the ultimate precision in the determination of the Higgs trilinear
coupling $\lambda$ requires the full integrated luminosity
of the HL-LHC, $\mathcal{L}=3$ ab$^{-1}$,
and will rely on the combination of all relevant final states.

In addition to the gluon-fusion production mode, Higgs pairs
can also be produced in the vector-boson fusion
channel $hhjj$~\cite{Contino:2010mh,Dolan:2013rja,Dolan:2015zja,
  Brooijmans:2014eja},
the associated production modes
$hhW$ and $hhZ$~\cite{Barger:1988jk,baglio,Cao:2015oxx}
(also known as Higgs-Strahlung),
and also in association
with top quark pairs $hht\bar{t}$~\cite{Englert:2014uqa}.
%
All these channels are challenging due to the small production
rates: at 14 TeV, the inclusive total cross-sections are
2.0 fb for VBF $hhjj$~\cite{Liu-Sheng:2014gxa},
0.5 fb for $W(Z)hh$~\cite{baglio}
and 1.0 for $hht\bar{t}$~\cite{Englert:2014uqa}.
%
The interplay between  kinematic
distributions for the
extraction of $\lambda$ from the measured
cross-sections, and the role of the associated theoretical
uncertainties, have been intensely scrutinized
recently~\cite{Slawinska:2014vpa,Chen:2014xra,Goertz:2013kp,
  Frederix:2014hta,Dawson:2015oha,Maltoni:2014eza,Maierhofer:2013sha,Grigo:2013rya,Grigo:2014jma}.
%

While the SM production rates for Higgs
pairs are small, these can be substantially
enhanced in a variety of BSM scenarios.
%
Feasibility studies of Higgs pair production in New Physics
models have been performed in a number of different frameworks,
including Effective Field
Theories (EFTs) with higher-dimensional
operators and anomalous 
Higgs couplings~\cite{Nishiwaki:2013cma,Dall'Osso:2015aia,Azatov:2015oxa,Liu:2014rba,Goertz:2014qta,He:2015spf,Grober:2015cwa}, resonant production
in models such as extra dimensions~\cite{Gouzevitch:2013qca,Cooper:2013kia,No:2013wsa,Wen-Juan:2015gqg}, and Supersymmetry and
Two Higgs Doublet models (2HDMs)~\cite{Belyaev:1999kk,Han:2013sga,Hespel:2014sla,Wu:2015nba,Cao:2014kya,Ellwanger:2013ova}.
%
Since BSM dynamics modify
the kinematic distributions of the Higgs decay products, for
instance boosting the di-Higgs system,
different analysis strategies  might be required for BSM
Higgs pair searches as compared to SM measurements.

Searches for the production of Higgs pairs
have already been performed with 8 TeV Run I data
by ATLAS in the $b\bar{b}b\bar{b}$~\cite{Aad:2015uka}
and $b\bar{b}\gamma\gamma$~\cite{Aad:2014yja} final states,
and by
CMS in the same $b\bar{b}b\bar{b}$~\cite{Khachatryan:2015yea}
and $b\bar{b}\gamma\gamma$~\cite{Chatrchyan:2011wt} final
states.
%
%
In addition, ATLAS has presented~\cite{Aad:2015xja} a combination
of its di-Higgs searches in the $bb\tau\tau,$
$\gamma\gamma WW^*$, $\gamma\gamma bb$, and $bbbb$ final states.
%
Many other exotic searches involves Higgs pairs in the final
state, such as the recent
searches for heavy Higgs bosons $H$~\cite{Khachatryan:2015tha}.

In the context of SM production,
the main advantage of the $b\bar{b}b\bar{b}$ final state is the
enhancement of the signal yield
from the large branching fraction of Higgs bosons into bottom-anti-bottom
pairs, ${\rm BR}\lp H\to b\bar{b}\rp\simeq 0.57$~\cite{Dittmaier:2012vm}.
%
However, a measurement in this channel
needs to deal with an overwhelming QCD multi-jet background.
%
Recent studies of Higgs pair production in this
final state~\cite{Wardrope:2014kya,deLima:2014dta}
estimate that, for an integrated
luminosity of
$\mathcal{L}=3$ ab$^{-1}$,
a signal significance of around $S/\sqrt{B}\sim 2.0$ could be obtained.
%
While irreducible backgrounds, such as $4b$ and
$t\bar{t}$, are included in these analysis, the contribution
from reducible processes, such as $bbjj$ and
$jjjj$, which can contribute to the signal yield when 
light jets are mis-identified as $b$-jets,
are not accounted for.
%
As will show, due to both
selection effects and $b$-quark radiation in the
parton shower, the
contribution of the $2b2j$ process is comparable to
the irreducible $4b$ component.

In this work, we revisit the feasibility of SM Higgs pair production in
gluon-fusion
in the $b\bar{b}b\bar{b}$ final state at the LHC.
%
 Our strategy is based on a combination of traditional cut-based
 methods with multivariate analysis (MVA).
  %
  We account for  all relevant
  backgrounds, including the contribution from mis-identified
  light and charm jets.
  %
  We also assess the robustness of our analysis strategy in
  an environment with high pile-up (PU).
  %
  We find an overall signal significance of $S/\sqrt{B}\simeq 4.0$
  for a total integrated
  luminosity of $\mathcal{L}=3$ ab$^{-1}$,
   indicating
  that the $b\bar{b}b\bar{b}$ channel
alone should allow for the observation of double Higgs production
  at the HL-LHC.
%

The structure of this paper is the following.
%
In Sect.~\ref{mcgeneration} we present the modeling of the signal
and background processes with Monte Carlo event generators.
%
In Sect.~\ref{sec:analysis}
we introduce our analysis strategy, and in particular
the classification of individual events into
different categories according to their topology.
%
The results the cut-based analysis
are then presented in Sect.~\ref{sec:results}.
%
In Sect.~\ref{sec:mva} we illustrate the enhancement of signal
significance using multivariate techniques, and
assess the robustness of our results in a high-PU environment.
%
In Sect.~\ref{sec:conclusions} we conclude and outline
future studies to extend our results to estimate the accuracy
in the determination of trilinear coupling $\lambda$, as well as
to provide
 constraints in
BSM scenarios.

>>>>

The measurement of double Higgs production will be one of the central
physics goals of the LHC program in its recently started high-energy
phase, as well as for its future high-luminosity upgrade (HL-LHC)
which aims to accumulate a total integrated
luminosity of 3 ab$^{-1}$~\cite{ATLAS:2013hta,CMS:2013xfa}.
%
Higgs pair production is directly sensitive to the
Higgs trilinear coupling $\lambda$~\cite{baglio} and 
provides crucial
information on the electroweak symmetry breaking mechanism.
%
It also probes the underlying strength of the Higgs interactions
at high energies, and can be used to test the composite nature of the 
Higgs boson~\cite{Giudice:2007fh,Contino:2010mh}.
%
While Standard Model (SM) cross-sections are small,
many Beyond the SM (BSM)
scenarios predict enhanced rates for double Higgs production, therefore searches have already been performed by ATLAS and CMS with Run I data~\cite{Aad:2015xja,Aad:2015uka,Aad:2014yja,Khachatryan:2015yea,Chatrchyan:2011wt}
and will continue at Run II.
%
The study of Higgs pair production will also be relevant to
any future high-energy 
collider, either at a 100 TeV circular machine~\cite{Arkani-Hamed:2015vfh,Barr:2014sga,Papaefstathiou:2015iba,Azatov:2015oxa} or at
a linear or circular electron-positron collider~\cite{Contino:2013gna}.

Analogously to single Higgs production~\cite{Dittmaier:2012vm}, 
in the SM the dominant mechanism for the production of a pair of
Higgs bosons at the LHC is 
gluon fusion (see Refs.~\cite{baglio,Frederix:2014hta} and
references therein).
%
For a center-of-mass energy of $\sqrt{s} = 14\,$TeV, the
next-to-next-to-leading order (NNLO)
total cross section is approximately $40\,$fb~\cite{deFlorian:2013jea},
which is increased by a further few percent once
next-to-next-to-leading logarithmic
(NNLL) corrections
are accounted for~\cite{deFlorian:2015moa}.
%
Feasibility studies in the case of a SM-like Higgs boson
in the gluon-fusion channel
at the LHC have been performed for different final states, including
$b\bar b\gamma\gamma$~\cite{Baur:2003gp,Barger:2013jfa},
$b\bar{b}\tau^+\tau^-$~\cite{Baur:2003gpa,Barr:2013tda,Dolan:2012rv,Dolan:2013rja},
$b\bar{b}W^+W^-$~\cite{Dolan:2012rv,Papaefstathiou:2012qe} and
$b\bar{b}b\bar{b}$~\cite{Baur:2003gpa,Dolan:2012rv,Wardrope:2014kya,deLima:2014dta,Barger:2013jfa}.
%
While these studies differ in their quantitative conclusions,
a consistent picture emerges 
that the ultimate precision in the determination of the Higgs trilinear
coupling $\lambda$ requires the full integrated luminosity
of the HL-LHC, $\mathcal{L}=3$ ab$^{-1}$,
and will rely on the combination of all relevant final states.

In addition to the gluon-fusion production mode, Higgs pairs
can also be produced in the vector-boson fusion
channel $hhjj$~\cite{Contino:2010mh,Dolan:2013rja,Dolan:2015zja,
  Brooijmans:2014eja},
the associated production modes
$hhW$ and $hhZ$~\cite{Barger:1988jk,baglio,Cao:2015oxx}
(also known as Higgs-Strahlung),
and also in association
with top quark pairs $hht\bar{t}$~\cite{Englert:2014uqa}.
%
All of these channels are challenging due to the small production
rates: at 14 TeV, the inclusive total cross-sections are
2.0 fb for VBF $hhjj$~\cite{Liu-Sheng:2014gxa},
0.5 fb for $W(Z)hh$~\cite{baglio}
and 1.0 for $hht\bar{t}$~\cite{Englert:2014uqa}.
%
The interplay between  kinematic
distributions for the
extraction of $\lambda$ from the measured
cross-sections, and the role of the associated theoretical
uncertainties, have been intensely scrutinized
recently~\cite{Slawinska:2014vpa,Chen:2014xra,Goertz:2013kp,
  Frederix:2014hta,Dawson:2015oha,Maltoni:2014eza,Maierhofer:2013sha,Grigo:2013rya,Grigo:2014jma}.
%

While the SM production rates for Higgs
pairs are small, they can be substantially
enhanced in a variety of BSM scenarios.
%
Feasibility studies of Higgs pair production in New Physics
models have been performed in a number of different frameworks,
including Effective Field
Theories (EFTs) with higher-dimensional
operators and anomalous 
Higgs couplings~\cite{Nishiwaki:2013cma,Dall'Osso:2015aia,Azatov:2015oxa,Liu:2014rba,Goertz:2014qta,He:2015spf,Grober:2015cwa}, resonant production
in models such as extra dimensions~\cite{Gouzevitch:2013qca,Cooper:2013kia,No:2013wsa,Wen-Juan:2015gqg}, and Supersymmetry and
Two Higgs Doublet models (2HDMs)~\cite{Belyaev:1999kk,Han:2013sga,Hespel:2014sla,Wu:2015nba,Cao:2014kya,Ellwanger:2013ova}.
%
Since BSM dynamics modify
the kinematic distributions of the Higgs decay products, for
instance boosting the di-Higgs system,
different analysis strategies  might be required for BSM
Higgs pair searches as compared to SM measurements.

Searches for the production of Higgs pairs
have already been performed with 8 TeV Run I data
by ATLAS in the $b\bar{b}b\bar{b}$~\cite{Aad:2015uka}
and $b\bar{b}\gamma\gamma$~\cite{Aad:2014yja} final states,
and by
CMS in the same $b\bar{b}b\bar{b}$~\cite{Khachatryan:2015yea}
and $b\bar{b}\gamma\gamma$~\cite{Chatrchyan:2011wt} final
states.
%
%
In addition, ATLAS has presented~\cite{Aad:2015xja} a combination
of its di-Higgs searches in the $bb\tau\tau,$
$\gamma\gamma WW^*$, $\gamma\gamma bb$, and $bbbb$ final states.
%
Many other exotic searches involve Higgs pairs in the final
state, such as the recent
search for heavy Higgs bosons $H$~\cite{Khachatryan:2015tha}.

In the context of SM production,
the main advantage of the $b\bar{b}b\bar{b}$ final state is the
enhancement of the signal yield
from the large branching fraction of Higgs bosons into bottom-anti-bottom
pairs, ${\rm BR}\lp H\to b\bar{b}\rp\simeq 0.57$~\cite{Dittmaier:2012vm}.
%
However a measurement in this channel
needs to deal with an overwhelming QCD multi-jet background.
%
Recent studies of Higgs pair production in this
final state~\cite{Wardrope:2014kya,deLima:2014dta}
estimate that, for an integrated
luminosity of
$\mathcal{L}=3$ ab$^{-1}$,
a signal significance of around $S/\sqrt{B}\sim 2.0$ could be obtained.
%
While irreducible backgrounds, such as $4b$ and
$t\bar{t}$, are included in these analysis, the contribution
from reducible processes, such as $bbjj$ and
$jjjj$, which can contribute to the signal yield when 
light jets are mis-identified as $b$-jets,
are not accounted for.
%
As we will show, due to both
selection effects and $b$-quark radiation in the
parton shower, the
contribution of the $2b2j$ process is as significant as
the irreducible $4b$ component.

In this work, we revisit the feasibility of SM Higgs pair production by
gluon-fusion
in the $b\bar{b}b\bar{b}$ final state at the LHC.
%
 Our strategy is based upon a a combination of traditional cut-based
 methods and multivariate analysis (MVA).
  %
  We account for  all relevant
  backgrounds, including the contribution from mis-identified
  light and charm jets.
  %
  We also assess the robustness of our analysis strategy in
  an environment including pile-up (PU).
  %
  We find an overall signal significance of $S/\sqrt{B}\simeq 4.0$
  for a total integrated
  luminosity of $\mathcal{L}=3$ ab$^{-1}$,
   indicating
  that the $b\bar{b}b\bar{b}$ channel
alone may allow for the observation of double Higgs production
  at the HL-LHC.
%

The structure of this paper proceeds as follows.
%
In Sect.~\ref{mcgeneration} we present the modeling of the signal
and background processes with Monte Carlo event generators.
%
In Sect.~\ref{sec:analysis}
we introduce our analysis strategy, in particular
the classification of individual events into
different categories according to their topology.
%
Results of the cut-based analysis
are then presented in Sect.~\ref{sec:results}.
%
In Sect.~\ref{sec:mva} we illustrate the enhancement of signal
significance using multivariate techniques, and
assess the robustness of our results against the effects of PU.
%
In Sect.~\ref{sec:conclusions} we conclude and outline
future studies to extend our results to estimate the accuracy
in the determination of trilinear coupling $\lambda$, as well as
to provide
 constraints in
BSM scenarios.
